Synthesizing Symmetric Lenses

### 2018-06-30 papers/SymmetricLenses/main.tex
edits in progress

  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|>{\columncolor{lightyellow}}c|>{\columncolor{lightgreen}}c|}
      \hline
      \texttt{company.mgmt} data file & \texttt{company.hr} data file \\
      \hline
      \listingA & \listingB \\
      \hline
      \cellcolor{yellow}\texttt{company.mgmt} regular expression & 
      \cellcolor{green}\texttt{company.hr}  regular expression\\
      \hline
      \cellcolor{yellow}\listingC & \cellcolor{green}\listingD \\
      \hline
    \end{tabular}
  }
  \caption{Example data files and corresponding regular expressions
  used by management and HR, respectively, to represent employee salaries and health
  insurance companies.  
   \ksf{fix colors so regular expressions are just slightly darker
  shade than data file.}}  
  \label{fig:minimized-representations}
\end{figure}

@@

We use regular expressions to denote the types of data formats, and we
impose a typing discipline on our lenses.  Specifically, if 
$\Regex$ and $\RegexAlt$ are regular expressions, then the notation
\lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$} indicates that $\Lens$
is a simple symmetric lens between formats described by $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. 
When clear from context, we use the term \textit{lens} to denote a
simple symmetric lens.

@@

To alleviate these difficulties, we wish to synthesize symmetric lenses---given
a pair of regular expressions and a set of input-output examples, we want to
find a symmetric lens expression that is typed between the regular expression
pair, and satisfies the input/output examples. In this example, we wish to
synthesize a lens from \lstinline{emp_salaries} and \lstinline{emp_insurance},
using the strings in Figure~\ref{fig:minimized-representations} as an
input-output example for \CreateR.

However, our simple symmetric lens combinators permit many well-typed programs
between two regular expressions. For example, one possible lens between
\lstinline{emp_salaries} and \lstinline{emp_insurance} is the one in
Figure~\ref{fig:example_lens}, but the lens
%
\begin{lstlisting}
disc(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is also a valid lens (where \lstinline{rep($\Regex$)} is a string in the
language of $\Regex$). Furthermore, adding the synchronized files from
Figure~\ref{fig:minimized-representations} as an input-output example helps
little. The default strings given to \Disconnect merely need to be the strings
in the input-output examples, and the lens will still satisfy the specification.
While the disconnect lens is prohibited by two examples, or by a carefully
chosen input-output example on \PutR or \PutL, this problem continues presenting
itself for the sublenses, and is highlighted even more on larger formats. While
we could merely ask programmers to provide large, tight specifications for their
synthesis tasks, we would prefer the algorithm to choose the right lens
automatically. Instead of merely finding \emph{any} satisfying lens, we wish to
synthesize the \emph{best} satisfying lens.

But what is the best satisfying lens? Here, we take the view that the best satisfying lens is the
one that aligns as much data as is permitted by the type system and the
examples. More formally, we wish to minimize the expected number of bits
required to recover data in one format from a synchronized string in the other.
The expected number of bits is a well known concept in information theory: the
\emph{entropy} of that data source.

The entropy of a data source is based on the probability distribution of that
data source. Given a set $S$, and a probability distribution over that set $P
\OfType S \rightarrow \Reals$, the entropy of the $(S,P)$ pair is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$. While we have ways of expressing sets of strings (with
regular expressions), we do not have probability distributions. Here, we
express these probability distributions jointly with the language through
\emph{stochastic regular expressions}, regular expressions annotated with
probabilities. Our algorithm guesses a reasonable probability distribution from
the syntax, though if users manually provided stochastic regular expressions, or
a large suite of data in the formats through which we could learn the annotations,
we could instead use that.

With sets and probability distributions, we can find the expected number of bits
required to recover one string from a synchronized string. This expected value
is the conditional entropy of the stochastic regular expression ($\Regex$) given
another stochastic regular expression ($\RegexAlt$), and a lens that shows which
strings are synchronized with which ($\Lens$). We aim to find the lens that
minimizes the sum of that entropy with the conditional entropy of $\RegexAlt$
given $\Regex$ and $\Lens$.

>>>>

  \centering
  \resizebox{\columnwidth}{!}{%
    \begin{tabular}{|>{\columncolor{lightyellow}}c|>{\columncolor{lightgreen}}c|}
      \hline
      \texttt{company.mgmt} data file & \texttt{company.hr} data file \\
      \hline
      \listingA & \listingB \\
      \hline
      \cellcolor{vlightyellow}\texttt{company.mgmt} regular expression & 
      \cellcolor{vlightgreen}\texttt{company.hr}  regular expression\\
      \hline
      \cellcolor{vlightyellow}\listingC & \cellcolor{vlightgreen}\listingD \\
      \hline
    \end{tabular}
  }
  \caption{Example data files and corresponding regular expressions
  used by management and HR, respectively, to represent employee salaries and health
  insurance companies.  }  
  \label{fig:minimized-representations}
\end{figure}

@@

We use regular expressions to denote the types of data formats, and we
impose a typing discipline on our lenses.  Specifically, if 
$\Regex$ and $\RegexAlt$ are regular expressions, then the notation
\lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$} indicates that $\Lens$
is a simple symmetric lens between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. 
When clear from context, we use the term \textit{lens} to denote a
simple symmetric lens.

@@

To alleviate these difficulties, we wish to synthesize symmetric lenses---given
a pair of regular expressions and a set of input-output examples, we want to
find a symmetric lens expression that is typed between the regular expression
pair, and satisfies the input/output examples. We call such a lens
a \emph{satisfying lens}.
In this example, we wish to
synthesize a lens from \lstinline{emp_salaries} and \lstinline{emp_insurance},
using the example data in Figure~\ref{fig:minimized-representations} as an
input-output example from which to draw the required defaut values for 
\Disconnect lenses.

A challenge is that the simple symmetric lens combinators permit many
well-typed lenses
between a given pair of regular expressions. For example, 
Figure~\ref{fig:example_lens} gives one possible lens between regular
expressions  \lstinline{emp_salaries} and \lstinline{emp_insurance},
but 
%
\begin{lstlisting}
disc(emp_salaries,emp_insurance,rep(emp_salaries),rep(emp_insurance))
\end{lstlisting}
%
is another valid lens (where \lstinline{rep($\Regex$)} is a string in the
language of $\Regex$). Furthermore, adding the synchronized files from
Figure~\ref{fig:minimized-representations} as an input-output example helps
little. The default strings given to \Disconnect merely need to be the strings
in the input-output examples, and the lens will still satisfy the specification.
While the disconnect lens is prohibited by two examples, or by a carefully
chosen input-output example on \PutR or \PutL, this problem continues presenting
itself for the sublenses, and is highlighted even more on larger
formats. 
\ksf{I don't understand the above.  We shouldn't explain the issue in
terms of \PutR and \PutL as we are focused on the combinators rather
than the raw functions.}
While
we could ask programmers to provide detailed and precise specifications for their
synthesis tasks, we would prefer the algorithm to choose the right lens
automatically. Instead of merely finding \emph{any} satisfying lens, we wish to
synthesize the \emph{best} satisfying lens.

But what is the best satisfying lens? We propose that the
best satisfying lens is the satisyfing lens that connects as much data
as possible between the two formats. More formally, we
propose that the best satisfying lens is the satisfying lens that 
minimizes the expected number of bits required to recover data in
one format from synchronized data in the other.  
In more detail, we say that two strings $s$ and $t$
are \emph{synchronized according to lens $l$} if 
$l.\PutR\ s\ t = t$ and $l.\PutL\ t\ s = s$ where the notation $l.\PutR$
extracts the \PutR function from simple symmetric lens $l$; similarly
for \PutL.
We say that we can \emph{recover} $s$ from $t$ using bits $b$ and lens
$l$ if we can reconstruct $s$ from $t$, $b$, and $l$.  For example,
given the \IdentityLens{} lens, we can recover $s$ from $t$ using
no bits because in this case, $s$ is just $t$.  In contrast, given
the \Disconnect{} lens, we need enough bits to fully encode $s$ in
order to recover it from $t$ because $t$ contains no useful
information.  This calculation shows that if both \IdentityLens{}
and \Disconnect{} are satisfying lenses, then \IdentityLens{} is
better.

The expected number of bits required to recover some data is a
well-known concept in information theory: it is the \emph{entropy} of
the data. Calculating the 
entropy of some data requires a probability distribution over the
space of possible values the data can take on.  Specifically,
given a set $S$ and a probability distribution over that set $P
\OfType S \rightarrow \Reals$, the entropy of the $(S,P)$ pair is $-\Sigma_{s\in
  S}P(s)\Log_2P(s)$.

In our setting, we already have a way of expressing sets of data:
regular expressions.  What we need is a way to express probability
distributions over that data.
To that end, we adopt \emph{stochastic regular expressions}\ksf{is
there a citation?}, 
which are regular expressions where each operator is annotated with a
probability.  Stochastic regular expressions
allow us to jointly specify the allowable data and give a probability
distribution over it.  From this information, we can calculate the
entropy of the stochastic regular expression.

Given a lens $\Lens$ between two stochastic regular expressions $\Regex$ and
$\RegexAlt$, we can calculate the expected number of bits required
to recover a string $s$ in $\LanguageOf{\Regex}$ from a synchronized
string $t$ in $\LanguageOf{\RegexAlt}$. 
This expected value is just the conditional entropy of the $\Regex$ given
$\RegexAlt$ and $\Lens$. 
To ensure symmetry, we seek to find the lens that
minimizes the sum of that entropy with the conditional entropy of $\RegexAlt$
given $\Regex$ and $\Lens$.

\ksf{This piece might better go somewhere deeper in the paper.}
Our algorithm guesses a reasonable probability distribution from
the syntax, though if users manually provided stochastic regular expressions, or
a large suite of data in the formats through which we could learn the annotations,
we could instead use that.
